Parallelization of Push-based System for Molecular Simulation Data Analysis with GPU

### 2016-02-15 paper/paperSource.tex
automatic triggered changes

\emph{\textbf{GPU execution}}. From hardware standpoint, it is important to understand that every GPU device consists of SM (streaming multiprocessors), and each SM has thousands of registers, which can be partitioned among execution threads. Having this in mind, since this is a pure parallel processing power, the goal is to increase hardware throughput and keep highest utilization to exploit every register given. CUDA is an amazing tool that tries to help developers map this hardware capabilities with the actual application layer. In CUDA we have grids, blocks, and threads, where grid is 3 dimensional unit where each cell is a block, and the same relationship applies between blocks and threads\footnote{This explanation doesn't include memory setup in GPU devices, but it is covered in further sections}. Originally, this mapping was perfect for solving image processing problems or other graphics computation, since dimensions and such division (e.g. by pixels to threads) is very inherent. As a language feature, in CUDA we have to define such functions called 'kernel' function. Essentially, it is the function where we are supposed to map every execution thread with data and perform the actual execution, this approach is widely known as SIMD\footnote{SIMD - Single Instruction Multiple Data}. As a convenience we have separated one-body functions with multiple-body functions into separate kernels.

\emph{\textbf{One-body queries kernel}}.
Commonly used CUDA array reduction\cite{gpureduction} makes the implementation of all the one-body queries obvious. Figure \ref{fg:reduction} nicely represents general idea of the data mapping, where each cell of the vector can be associated with a thread of execution in the kernel. Having an array of atom data structures, we reduce the array accumulating all the needed data from each atom's attributes. We don't expect much performance improvement from coalesced memory access, since the data structure for atom can eliminate any gain of spacial prefetch, though for specific types of queries it could be possible to aggregate certain attributes into merged arrays and take an advantage. In this case, the solution should be simple in order to keep it easily extendable for other simple one-body queries. We fetch all the atoms from GPU global memory, since we use references to atoms once. In this case, allocating shared memory and copying it over would complicate the implementation and possibly create additional overhead for one time reference usage.

\emph{\textbf{Two-body queries kernel}}.
In general, queries involving two-body functions are a
bit more complex and cannot provide the final result in a single
data read if a straightforward method is used for their computation.
However, in the proposed system we have incorporated a data
structure and an algorithm for the SDH (also RDF) from our previous
work that is suitable for push-based type of system. In this
subsection we give a brief description of the data structure (DM)
and the algorithm (DM-SDH) designed in~\cite{ytu:icde09, EDBT12} and
implemented in the system proposed in this paper. For more detailed
information, please refer to our previously published work on this
topic~\cite{ytu:icde09, EDBT12}

\emph{\textbf{The data structure.}} The simulation data space is
represented by a conceptual data structure we named Density Map
(DM). The density map splits the simulation space into a grid of
equal size regions (or cells). The cells are cubes in 3D and squares
in 2D\footnote{In this paper, we focus only on the 2D data to
elaborate and illustrate the proposed ideas.} Resolution of a
density map is the reciprocal of the cell size in that density map.
In order to generate higher resolution density map, we split each
cell of the current resolution's grid into four smaller cells of
equal size. This design allows us to use a region
quad-tree~\cite{orenstein:ipl82} to organize density maps of the
same data but with different resolutions. So, essentially, a node in
the quad-tree represents a single cell from the DM. Therefore, a
density map of a certain resolution basically is the set of all
nodes of one level of the tree. Each of the tree nodes records the
cell's location in the density map (coordinates of corner points) as
well as the number of particles in each cell. We name the afore
describe tree the Density-Map tree (DM-tree).

\begin{figure}
 \centerline{ \includegraphics[width=0.65\columnwidth]{images/reduction} }
 \caption{Array reduction using coalesced memory access\cite{gpureduction}}
 \label{fg:reduction}
\end{figure}

\emph{\textbf{The algorithm.}} The essential part of the DM-SDH
algorithm is a procedure named {\sc ResolveTwoCells}. The input to
this procedure are two cells from the density map (e.g., $A$ and $B$
in Fig.~\ref{fg:resolve}). It computes, in constant time, the
minimum and maximum distance between the two cells. A pair of cells
is ~\emph{resolvable} if both the min and max distance between them
fall into the same SDH bucket $i$. If that is the case, the distance
count of that bucket is being increase by $n_{A}n_{B}$ ($n_{A}$ and
$n_{B}$ are the number of particles in cell $A$ and $B$,
respectively). Otherwise, the cells are non-resolvable and we
either:
\begin{itemize}
\item[(1)] Go to the next density map with higher resolution and resolve all children of $A$
with those of $B$, or
\item[(2)] If leaf-level has been reached: compute every distance between particles of $A$ and $B$ and update the
histogram accordingly.
\end{itemize}

In order to generate the complete SDH, the {\sc ResolveTwoCells}
procedure is executed for all pairs of cells for a given density map
$DM_k$ and the algorithm would recursively call the procedure
(action (1) above) until leaf-level has been reached (action (2)
above).

On top of the aforementioned DM-SDH algorithm, we have also
incorporated two approximate SDH algorithms (ADM-SDH), introduced
and described in \cite{EDBT12}, and \cite{TKDE12}. These approximate
algorithms are substantially faster than the the brute-force
algorithm and also than the DM-SDH algorithm as they take advantage
of some heuristic. For more details on the ADM-SDH algorithms please
see the aforementioned work.

@@

\subsection{Working of the system}

@@

Fig.~\ref{fg:system} depicts the flow of the system.

\begin{figure}
 \centerline{ \includegraphics[width=.8\columnwidth]{images/system.eps} }
 \caption{Push-based system flow}
 \label{fg:system}
\end{figure}

>>>>

\emph{\textbf{GPU execution}}. From hardware standpoint, it is important to understand that every GPU device consists of SM (streaming multiprocessors), and each SM has thousands of registers, which can be partitioned among execution threads. Having this in mind, since this is a pure parallel processing power, the goal is to increase hardware throughput and keep highest utilization to exploit every register given. CUDA is an amazing tool that tries to help developers map this hardware capabilities with the actual application layer. In CUDA we have grids, blocks, and threads, where grid is 3 dimensional unit where each cell is a block, and the same relationship applies between blocks and threads\footnote{This explanation doesn't include memory setup in GPU devices, but it is covered in further sections}. Originally, this mapping was perfect for solving image processing problems or other graphics computation, since dimensions and such division (e.g. by pixels to threads) is very inherent. As a language feature, in CUDA we have to define such a function called 'kernel' function. Essentially, it is the function where we are supposed to map every execution thread with data and perform the actual execution, this approach is widely known as SIMD\footnote{SIMD - Single Instruction Multiple Data}. As a convenience we have separated one-body functions with multiple-body functions into separate kernels.

\emph{\textbf{One-body queries kernel}}.
Commonly used CUDA array reduction\cite{gpureduction} makes the implementation of all the one-body queries obvious. Figure \ref{fg:reduction} nicely represents general idea of the data mapping, where each cell of the vector can be associated with a thread of execution in the kernel. Having an array of atom data structures, we reduce the array accumulating all the needed data from each atom's attributes. We don't expect much performance improvement from coalesced memory access, since the data structure for atom can eliminate any gain of coalesced memory access, though for specific types of queries it could be possible to aggregate certain attributes into merged arrays and take an advantage of it. In this case, the solution should be simple in order to keep it easily extendable for other simple one-body queries for a scientist to program. We fetch all the atoms from GPU global memory, since we use references to atoms once. In this case, allocating shared memory and copying it over would complicate the implementation and possibly create additional overhead for one time reference usage. Finally, memory transportation and data consumption are implemented for user, thus, he only needs to extend the queries module which consequently can be turned into plugins model.

\emph{\textbf{Two-body queries kernel}}.
SDH (or RDF) is the algorithm that requires special treatment in our design. The point, as mentioned earlier, is that this function requires more than one iteration of the given set of atoms, but rather all combinations of pairs in a given frame, thus, it is not linear complexity. An implementation of this type of query might be very non-intuitive because of the nature of GPUs. Another issue is that since to take an advantage on this type of computation, before writing an implementation for it, a user must have a good understanding of GPU architecture which makes the system less user-friendly. At this point, it is more important to achieve good performance improvement results and then take care of ease of system extensions. As we already mentioned, GPU threads execute in parallel and consume certain amount of work followed by best try of all hardware capacities utilization, but, unfortunately, it is also important to take care of equal workload for every thread independently, so, that utilization is kept high. For this reason, we've come up with a solution that should increase performance of the computation which will be described later.

\begin{figure}
 \centerline{ \includegraphics[width=0.65\columnwidth]{images/reduction} }
 \caption{Array reduction example\cite{gpureduction}}
 \label{fg:reduction}
\end{figure}

\begin{figure}
 \centerline{ \includegraphics[width=0.65\columnwidth]{images/overallarchitecture} }
 \caption{Overall system architecture}
 \label{fg:overallarchitecture}
\end{figure}

\emph{\textbf{Architecture overview.}}
This section is to emphasize general architecture of the system and summarize the work flow as well as the data flow. In total, we have flat files with ready flat values for each atom aggregated in frames coming into the program as data input using common disk I/O framework of an operating system. Then, once we have loaded a frame (which is considered to be a chunk of the entire data set) onto RAM of our machine, we copy over the entire frame onto the GPU device global memory. It is important that, at this point, we never write any values to simulation files or any other data related to simulation, thus, we don't have to care about any synchronization of incoming data. After this, we have two predefined kernel functions (which were explained before), where first kernel stands for one-body queries, and the second one is for two-body query SDH. Just for architecture clearance and possible future improvements, two kernels are executed in two different streams\footnote{In CUDA stream is another abstract layer of organizing execution threads. They are not guaranteed to be executed at the same time or after each other, GPU scheduling architecture is completely different from CPU having a warp to be single scheduling unit at maximum of 32 threads at a time} (You can see them on Figure\ref{fg:overallarchitecture}). In each kernel we apply certain heuristics to improve memory access using such architecture features of GPU as shared memory and blocking of kernel threads.

\emph{\textbf{GPU grid and block sizes.}} As it has already been mentioned, utilization of hardware on GPU device is important, because it is the key to performance improvement, and the higher utilization is, the bigger improvement is.
The approach we came up with is 1 execution thread per atom. For CUDA version 2.0 maximum number of threads per block is 1024. Blocks are 3 dimensional entities in a 3 dimensional grid. The less abstract unit is, the less overhead we have in communication between them, thus, we try to utilize smallest units more. For example, for 1000,000 atoms, we would allocate 1024 threads per block as a static hard coded value, since we usually have always more than 1024 atoms in simulations. One dimension of grid can take up to 64,000 blocks, thus, if N is bigger than $1024 * 64,000$, than the rest would be allocated to $grid.y$ and $grid.z$ respectively. We have not considered numbers bigger than GPU can allocate, which could be explored in future work. It is important to understand that even though there are abstract entities of grids and blocks, it is not real hardware capacities, but they can affect scheduling and performance consequently. Decision to go with atom to thread execution is because it is pretty convenient to keep high utilization of hardware and still not lose idea if we are about to incorporate more functions, which makes the system more user-friendly, though later on as a plugin we could add data loaded and transformer between RAM and GPU global memory, because there are a lot of custom ways to aggregate data and take an advantage of coalesced memory access.

@@

\subsection{Computation work flow}

@@
